Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-13 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Update plot, makefile, revise draft (partly 4.2 and 4.3)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2873 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 56644a1584918bfeda066dd1e229c8159def3c3a

\subsection{Evaluation and models}
Along with distributional trees and forests for circular data three other models have been applied
in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
the five years contained in the data set being used as testing data once after the models where 
learned on the remaining four years.

In order to compare the performances of the employed models a circular version of the continuous 
ranked probability score (CRPS) as introduced by \cite{Grimit+Gneiting+Berrocal:2006} is calculated.
Just as the linear version of the CRPS \citep[for more details see][]{Hersbach:2000} it measures the 
difference between an observation and the corresponding predicted distribution function in order 
to assess the probabilistic goodness of fit of the fitted model. Hence, the lower the CRPS value 
the better the performance. In order to adjust the general methodology of the CRPS to circular data 
\cite{Grimit+Gneiting+Berrocal:2006} employ an auxiliary function which measures the angular distance 
between two circular variables.

\fixme{condense evaluation}

The circular CRPS is evaluated out of sample and the resulting values are averaged over 
each observation contained in the testing data and the corresponding predicted distribution 
parameters of each of the models.

\fixme{comment on aggregation}

To measure the improvement of one of the applied methods over a climatology model (which will 
be described in detail subsequently) the skill score based on the circular 
CRPS of the assessed method is evaluated with the climatology model used as reference:
\begin{equation}
\text{CRPSS}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}

Three commonly used methods are selected as reference models in order to investigate whether 
distributional trees and forests can be applied as a reasonable alternative to already
existing approaches to modeling circular data, particularly in the field of meteorological forecasting.
The exact settings of the tree and forest models as well as a specification of the reference models
is provided in the remainder of this section.

\fixme{condense model descriptions}

\paragraph{Distributional Tree}
For the distributional tree as discussed in Section~\ref{sec:circtree} all covariates provided in the 
learning data are considered. The tree is obtained by employing the CTree algorithm with a minimal 
number of 2000 observations in each terminal node serving as a stopping criterion 
(argument \code{minbucket} of our implementation).

\paragraph{Distributional Forest}
Following the description in Section~\ref{sec:circforest} a distributional forest is constructed.
As for the distributional tree all available covariates are used to learn a distributional 
forest of 100 trees. Each of these trees is built by the CTree algorithm on a subsample 
containing a fraction of 30 percent of the original learning data. As it can be expected that 
the lagged response variable (as included as a covariate) has a strong influence on the 
response variable it is desirable to employ it in each tree of the forest. Therefore, all 
covariates are considered for each tree
(by setting the argument \code{mtry} of our implementation to the total number of covariates)
in order to guarantee for the lagged response to be available.
Since a high number of possible split points leads to a high computational effort that is 
particularly noteworthy for forest models the covariates are binned allowing for a maximum of
$50$ classes. In that way the number of split points that have to be considered is reduced
in a reasonably.
% control = disttree_control(nmax = c("yx" = Inf, "z" = 50)))
As stopping criteria the minimal number of observations to perform a split (\texttt{minsplit}) 
is set to 20, the minimal number of observations in each terminal node (\texttt{minbucket})
is set to 7, and the significance level for variable selection (\texttt{alpha}) is kept at its 
maximum value of 1 in order to build trees as large as possible as commonly done for forest models.

\paragraph{Persistence}
Assuming that weather conditions will remain the same after a certain period of time is the
basic idea of persistence in meteorology. While this simple way of making predictions can
be very effective particularly under stable atmospheric circumstances it can fail remarkably
as soon as any changes take place, especially in case of rapid changes.
Therefore, it heavily depends on the specific features and the variability of the variable
of interest whether this approach can provide overall reliable predictions.

@@

To establish a distributional model of persistence in this study a set of six observations
is considered to make a prediction for a selected day and time point. In particular,
the six previous time points are considered. In order to allow for a stronger influence of 
observations closer to the prediction time point exponential smoothing is employed, i.e.,
exponentially decaying weights are included with a smoothing factor of $0.5$ such that
in each prediction step the current prediction and the previous observation both have an 
equal influence rate of 50 percent. If a seventh or even earlier observations were considered 
as well their weights would be $<0.01$ and thus are not included anymore to simplify calculations.
For this selected set of weighted observations the distribution parameters of the von Mises 
distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
parametric model is specified allowing for probabilistic predictions and an evaluation of the 
circular CRPS.
As this approach only considers data from the previous six days a cross validation is not 
applicable, however, the same evaluation framework as for the other four models can be employed 
by obtaining predictions only for the testing data of the cross validation without using 
the corresponding learning data.

\paragraph{Climatology}
If there are no climatological changes within a certain number of years it can be expected that 
for a selected time point and day of the year weather conditions will always be the same over all 
years considered.
Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
meteorological variable of interest that can be very useful but its performance can also suffer 
dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.

@@

\paragraph{Generalized Linear Model}
Another commonly applied approach is to linearly regress the circular variable of interest on
its lagged values. For the here discussed application this corresponds to establishing a generalized
linear model with wind direction as the outcome variable and wind direction of the previous time point 
as covariate. 
Following \cite{Mulder+Klugkist:2017} the employed model plugs in a link function in order
to guarantee values on the interval $[-\pi,\pi]$ and is of the form
\begin{align}
dd = \beta_0 + g(\beta_1 \cdot dd_{lagged})
\end{align}
with $\beta_0$ being a circular intercept, $\beta_1$ the regression coefficient and 
$g$ the link function $g(x) = r \cdot \arctan(x)$ with an optional factor $r$. 
The intercept and the regression coefficient are estimated in a Bayesian framework where an 
MCMC algorithm is used to obtain posterior samples. Moreover, a constant concentration
parameter is fitted to the full learning sample.

\fixme{add references and further information on the estimation process within circGLM?}

\fixme{add information on applied implementation here (!) or in separate chapter?}

\subsection{Results}
\fixme{update results section}
This section provides a detailed analysis of the different proposed statistal models with focus
on their predictive performance. For a comprehensive comparison of the models, $1$\,hour and 
$3$\,h wind direction forecasts are evaluated at two airports, namely Innsbruck and Vienna, with
different meteorological site characteristics.

Figure~\ref{fig:boxplot_crpsraw} illustrates the CRPS values of the employed models with a lower 
CRPS indicating better performance. 
In all four settings it can be observed that on average the distributional forest provides the best 
fitting probabilistic predictions followed by the distributional tree. Especially in Vienna the 
persistence (hour) model is only slightly behind while the linear model and the climatology
show clearly higher CRPS values.

Comparing the performances over the different locations the narrower boxes in Vienna represent
a lower variation. This can be explained by more abrupt changes of wind directions in Innsbruck
due to the surrounding mountains channeling wind and in that way allowing only for a limited
number of possible wind directions (mainly the three wind regimes as discussed in the motivational
example in Section~\ref{sec:mot_ex}). For that reason, relatively stable models such as persistency
produce well fitting forecasts as long as wind directions remain in one of these possible states,
however, once a change takes place it can only be an abrupt and major shift leading to high
prediction errors as supported by the wide boxes of the persistence model in Innsbruck. In Vienna,
on the other hand, changes can also appear in a smaller extend as the geographic surroundings do not
constrain wind directions. This wider range of possible observations might also weaken the performance
of the climatology as expressed by the higher CRPS values in Vienna.

The impact of the different forecasting time windows of 1 and 3 hours can hardly be noticed, especially
as it does not seem to effect the order of which method performs best. But taking a closer look
at the CRPS values of the tree, forest and persistency reveals lower scores and hence a better 
performance for the 1-hour forecasts as it could have been expected due to increasing uncertainties
in a longer time window.

@@

With the aim of excluding any possible influence of the selection of testing and training data
the performances of the methods are also compared based on the CRPS skills score as illustrated in 
Figure~\ref{fig:boxplot_crpsskill}. Using the climatology as reference model the values on the y-axis
indicate the percentage of improvement of each model compared to the reference model. Overall
the same conclusions can be drawn as from Figure~\ref{fig:boxplot_crpsraw} with the distribution
forest leading to the highest overall improvement in model fit.

>>>>

\subsection{Models and evaluation}\label{sec:wind:models}
For a fair evaluation of circular regression trees and forest and to
investigate whether circular trees and forests can be applied as a reasonable
alternative to already existing approaches, three additional statistical models are
applied in this study to probabilistically forecast wind directions. Two
approaches are based on existing approaches in the field of meteorological
forecasting, one approach is a state-of-the-art model to forecast circular
variables.  The next paragraphs introduce the alternative approaches and
provide for all statistical models, including the circular tree and forest, the
model specification employed in this study.

\paragraph{Circular regression tree}
For the distributional tree as discussed in Section~\ref{sec:circtree} all covariates provided in the 
learning data are considered. The tree is obtained by employing the CTree algorithm with a minimal 
number of 2000 observations in each terminal node serving as a stopping criterion 
(argument \code{minbucket} of our implementation).

\paragraph{Circular regression forest}
Following the description in Section~\ref{sec:circforest} a distributional forest is constructed.
As for the distributional tree all available covariates are used to learn a distributional 
forest of 100 trees. Each of these trees is built by the CTree algorithm on a subsample 
containing a fraction of 30 percent of the original learning data. As it can be expected that 
the lagged response variable (as included as a covariate) has a strong influence on the 
response variable it is desirable to employ it in each tree of the forest. Therefore, all 
covariates are considered for each tree
(by setting the argument \code{mtry} of our implementation to the total number of covariates)
in order to guarantee for the lagged response to be available.
Since a high number of possible split points leads to a high computational effort that is 
particularly noteworthy for forest models the covariates are binned allowing for a maximum of
$50$ classes. In that way the number of split points that have to be considered is reduced
in a reasonably.
% control = disttree_control(nmax = c("yx" = Inf, "z" = 50)))
As stopping criteria the minimal number of observations to perform a split (\texttt{minsplit}) 
is set to 20, the minimal number of observations in each terminal node (\texttt{minbucket})
is set to 7, and the significance level for variable selection (\texttt{alpha}) is kept at its 
maximum value of 1 in order to build trees as large as possible as commonly done for forest models.

\paragraph{Persistency model}
Assuming that weather conditions will remain the same after a certain period of time is the
basic idea of persistence in meteorology. While this simple way of making predictions can
be very effective particularly under stable atmospheric circumstances it can fail remarkably
as soon as any changes take place, especially in case of rapid changes.
Therefore, it heavily depends on the specific features and the variability of the variable
of interest whether this approach can provide overall reliable predictions.

@@

To establish a distributional model of persistence in this study a set of six observations
is considered to make a prediction for a selected day and time point. In particular,
the six previous time points are considered. In order to allow for a stronger influence of 
observations closer to the prediction time point exponential smoothing is employed, i.e.,
exponentially decaying weights are included with a smoothing factor of $0.5$ such that
in each prediction step the current prediction and the previous observation both have an 
equal influence rate of 50 percent. If a seventh or even earlier observations were considered 
as well their weights would be $<0.01$ and thus are not included anymore to simplify calculations.
For this selected set of weighted observations the distribution parameters of the von Mises 
distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
parametric model is specified allowing for probabilistic predictions and an evaluation of the 
circular CRPS.
As this approach only considers data from the previous six days a cross validation is not 
applicable, however, the same evaluation framework as for the other four models can be employed 
by obtaining predictions only for the testing data of the cross validation without using 
the corresponding learning data.
\fixme{not done like this, always using the whole data set. wrong?}

\paragraph{Climatological model}
If there are no climatological changes within a certain number of years it can be expected that 
for a selected time point and day of the year weather conditions will always be the same over all 
years considered.
Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
meteorological variable of interest that can be very useful but its performance can also suffer 
dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.

@@

\paragraph{Generalized linear model}
Another commonly applied approach is to linearly regress the circular variable of interest on
its lagged values. For the here discussed application this corresponds to establishing a generalized
linear model with wind direction as the outcome variable and wind direction of the previous time point 
as covariate. 
Following \cite{Mulder+Klugkist:2017} the employed model plugs in a link function in order
to guarantee values on the interval $[-\pi,\pi]$ and is of the form
\begin{align}
dd = \beta_0 + g(\beta_1 \cdot dd_{lagged})
\end{align}
with $\beta_0$ being a circular intercept, $\beta_1$ the regression coefficient and 
$g$ the link function $g(x) = r \cdot \arctan(x)$ with an optional factor $r$. 
The intercept and the regression coefficient are estimated in a Bayesian framework where an 
MCMC algorithm is used to obtain posterior samples. Moreover, a constant concentration
parameter is fitted to the full learning sample.

%\fixme{add references and further information on the estimation process within circGLM?}

%\fixme{add information on applied implementation here (!) or in separate chapter?}

To compare the predictive performances of all proposed models a circular
analogue of the continuous ranked probability score (CRPS) as introduced by
\cite{Grimit+Gneiting+Berrocal:2006} is computed. Just as the linear version of
the CRPS \citep[for more details see][]{Hersbach:2000} it is a proper scoring
rule \citep{Gneiting+Raftery:2007} and measures the difference between an
observation and the corresponding predicted distribution function in order to
assess the probabilistic goodness of fit for the estimated model. Hence, the lower
the CRPS value the better the performance.  Contrary to the linear version, the
circular CRPS reduces not to the absolute error but to the angular distance
when the forecast is deterministic.

In addition to the raw CRPS, to assess differences in the
improvement of the various statistical models over the climatological model
used as a reference method, CRPS-based skill scores are computed according to
\begin{equation} 
  \text{CRPSS}_{\text{method}} = 1 -
\frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}

All scores presented in the next section are out-of-sample: For the persistency
model only time points prior to the time of interest are used, for all other
models a five-fold cross-validation is employed using up to four calendar years
for model training and the remaining single calendar year for validation. Due to the large sample
size of 24-hourly values per day over five years, some kind of temporal
aggregation is needed to ensure a correct visual comparison of the individual
methods. As for the employed models the variablilty of the predictive performance over the five years
is lower than over a single day or over a single year, CRPS and CRPS skill scores are
aggregrated over the respective five validation years; this yields 24 hourly
scores per 12 months averaged over the five validation years.

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistal models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (panels a,\,c) and
Vienna (panels b,\,d). The scores are aggregated over the $5$~validation years,
yielding yearly mean values for every hour per calendar month, with a lower
score indicating better performance.  For both stations at both forecast steps,
the circular regression forest overall provides the best predictive
performance, followed by the circular regression tree and, except for the
$3$\,h forecast at Innsbruck, the persistency model.  In comparison to the
circular tree and forest, for both stations and forecast steps, the
climatological model and the linear model show clearly higher CRPS scores and
hence a lower predictive performance.

The different site characteristics between the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}b,\,d) seem to have an effect on the absolute
level of the respective model performances and on their respective performance
variation. At Innsbruck, due to the surrounding mountains a limited number of
possible wind directions exists, namely the three wind regimes discussed for
Figure~\ref{fig:tree_ibk} in Section~\ref{sec:introduction}. Therefore, for
Innsbruck the wind direction remains rather constant in one of these possible
states, but once a change takes place it is mostly a major wind direction
shift.  Hence, due to the few wind regimes the rather unflexible climatological
and linear models perform relatively similar to the other models
(Figure~\ref{fig:boxplot_crpsraw}a,\,c). In addition, at Innsbruck the potential
high prediction errors in case of a paradigma change seem to lead to an higher
variation in the predictive performance for all models in comparison to Vienna;
this variation is especially high for the persistency model due to its strong
vulnerability to abrupt wind shifts. On the contrary, at Vienna the geographic
surroundings do not constrain any wind direction which leads to smaller and
less abrupt changes in the wind direction as well as less pronounced wind
regimes. This seems to weaken the predictive performance of the climatological
and linear models, and to reduce the performance variability for all models
(Figure~\ref{fig:boxplot_crpsraw}b,\,d).

The different forecast steps seem to have only a minor effect on the predictive
performance of the climatological model and the linear model for both stations.
However, for the persistency model, at both stations, higher scores for the
$3$\,h forecast (panels c,\,d) reveal a lower performance for longer lead times. This is
clearly due to at least 3-hourly instead of 1-hourly lagged response values
employed as covariates in the persistency model. Apparently, the circular
regression tree and forest can partially compensate the lower skill of
the lagged response values by other covariates, hence their predictive
perfomance only slightly decreases for the 3-hourly in comparison to the
1-hourly wind direction forecast. This compensation is especially evident for
Innsbruck, where the performance difference between the persistency model and
the tree-based methods significantly increases for the 3-hourly forecast, with
a significantly better performance for the circular tree and forest.

@@

In addition to the raw CRPS, Figure~\ref{fig:boxplot_crpsskill} provides CRPS
skill scores with the climatological model as a reference. Skill scores are in
percent, and positive values indicate an improvement in the predictive
performance over the reference. For all setups, the circular regression forest
has the highest skill scores with a mean performance gain of 13--25\% and 58--71\% for
Innsbruck and Vienna, respectively. As discussed for
Figure~\ref{fig:boxplot_crpsraw}, this improvement over the climatological
model is lower for Innsbruck due to the few number of predominat wind regimes
and hence a relatively good performance of the climatological model.
Additionally, Figure~\ref{fig:boxplot_crpsskill} shows that even if the
persistency model's performance is lower than the reference (panel c) the
tree-based models are significantly superior to the climatolical model used as
reference.
